---
---
---

```{r setup}
# install.packages("dplyr")
library(dplyr)
```

We will be practicing with dplyr functions using the starwars data set:

```{r}
head(starwars)
starwars
```

# Subset rows with `filter()`

`filter()` allows us to subset the **rows** of a dataset based on the values

```{r}
##Filter for rows with human Star Wars characters
filter(starwars, species == "Human")

##save to a new R object
humans <- filter(starwars, species == "Human")

##filter multiple columns at once
filter(starwars, species == "Human", sex == "male")

##returns characters with missing mass
filter(starwars, is.na(mass))

##filter for non humans
filter(starwars, species != "Human")
```

Comparison operators can be used when filtering:

-   `!` is **not**

-   `==` is **equals**

-   `&` is **and**

-   `|` is **or**

-   `<` is **less than**

-   `>` is **greater than**

-   `is.na()` returns `True` if values is *`NA`*

-   variable_name `%in%` vector_of_values, returns `True` if the value of variable_name is equal to one of the values in vector_of_values

```{r}
"yellow" %in% c("yellow", "fair", "tan") #returns true

##filter for characters with yellow, fair or tan skin
filter(starwars, skin_color %in% c("yellow", "fair", "tan"))

##filter for characters with brown or blond hair
filter(starwars, hair_color %in% c("brown", "blond"))
```

### GREPL

`grepl("characters_of_interest", variable_name)` returns `TRUE` if the value of `variable_name` contains the `characters_of_interest`.

```{r}
##filter for characters with hair color that contains "blond" (aka blond or blonde)
filter(starwars, grepl("blond", hair_color))

##filter for all characters with blue in their eyes
filter(starwars, grepl("blue", eye_color))
```

You can compare numeric variables and categorical variables with one and each other

-   nominal categorical variables (character vectors) are ranked alphabetically

-   ordinal categorical variables (factors) are ranked via built-in order

```{r}
## starwars characters with a name starting with letters before "D"
filter(starwars, name < "D")


```

`between(x, left, right)` returns true if left \<= x \<= right

### Practice

Find all starwars characters that are:

1.  missing a hair color

2.  are shorter than 100cm and are not droids

3.  were born between the years 100 and 200

4.  names do not start with "A"

```{r}
#1
filter(starwars, is.na(hair_color))

#2
filter(starwars, species != "Droid", height < 100)

#3
filter(starwars, between(birth_year, 100, 200))

#4
filter(starwars, !grepl("A", name))
```

# Ordering rows with `arrange()`

arrange() reorders rows based on the values of column(s)

```{r}
##reorder characters by name's alphabetical order
arrange(starwars, name)
#NOT ASCII, order: aAbBcCdD

##order by species and birth year
arrange(starwars, species)
#within species order by oldest to youngest(small birth year to big )
arrange(starwars, species, birth_year)
#order youngest to oldest(big birth year to small birth year)
arrange(starwars, species, desc(birth_year))
```

### Practice

1.  Arrange from tallest to shortest
2.  Arrange from shortest to tallest
3.  Alphabetize the Star Wars characters by their hair color then eye color

```{r}
arrange(starwars, desc(height))
arrange(starwars, height)
arrange(starwars, hair_color, eye_color)
```

# Subset columns with `select()`

```{r}
##shows the dataset with onlt the hair color, skin color and eye color
select(starwars, hair_color, skin_color, eye_color)
#save the dataset composed of those three rows
character_colors <- select(starwars, hair_color, skin_color, eye_color)
```

We can also select multiple columns in a row:

```{r}
##show all columns through name to eye_color
select(starwars, name:eye_color)#":" only works in select() function
```

We can also remove columns by name using a "-":

```{r}
select(starwars, -films, -vehicles, -starships)
select(starwars, -c(films, vehicles, starships))
select(starwars, -c(films:starships))
```

### Helpful functions

`everything()` is a shortcut for "everything else"

```{r}
select(starwars, species, everything()) #puts species row in front and then "everything else" after

##alphabatize columns
starwars_names <- names(starwars) #create a vector of column names
starwars_names <- sort(starwars_names) #sort star wars names
select(starwars, starwars_names)
```

`starts_with("abc")` matches column names that start with "abc"

`ends_with("abc")` matches column names that ends with "abc"

`contains("abc")` matches column names that contains "abc"

```{r}
#only columns that start with "h"
select(starwars, starts_with("h"))

#only columns that end with "d"
select(starwars, ends_with("d"))

#only columns that contain "or"
select(starwars, contains("or"))
```

### Practice

1.  Select each character's homeworld and species. What question might this subset of variables help answer?
2.  Brainstorm as many ways as possible to select hair_color, skin_color, eye_color and birth_year

```{r}
starwars

#1
select(starwars, homeworld, species)
#this helps answer the question related the population of species relative to each world

#2
select(starwars, hair_color:birth_year)
select(starwars, c(hair_color:birth_year))
select(starwars, hair_color, skin_color, eye_color, birth_year)
select(starwars, -c(name: mass), -c(species:homeworld))
select(starwars, contains("_"))
```

# Add new variables with `mutate()`

```{r}
##new dataset to play around with
smallwars <- select(starwars, height, mass, species)
smallwars

mutate(smallwars, 
       height_m = height/100)#add a new column(at the end of the dataset) named "height_m" where each variable is /100 from height column
```

We can also create multiple variables at once and we can refer to variables we just created

```{r}
smallwars <- select(starwars, height, mass, species)
mutate(smallwars,
       height_m = height/100,
       bmi = mass/height_m^2)

##only keeps the new variables:
transmute(smallwars,
       height_m = height/100,
       bmi = mass/height_m^2)
```

Summary statistics give one value for the entire column: `sum()`, `mean()`

```{r}
smallwars <- select(starwars, height, mass, species)

##sums the entire mass colum and creates a column where all values equal that sum
mutate(smallwars, sum = sum(mass, na.rm = T))
mutate(smallwars, sumhs = sum(height, mass, na.rm = T))#sums height and mass columns
```

Group by a variable using the `.by` parameter

```{r}
smallwars <- select(starwars, height, mass, species)
##takes average of mass per species
mutate(smallwars,
       species_mass = mean(mass, na.rm = T),
       .by = species) 
```

Change the type of a variable:

```{r}
smallwars <- select(starwars, height, mass, species)
mutate(smallwars, mass = as.character(mass)) #make mass column into characters
```

### Practice

1.  convert mass to pounds (1kg roughly 2.2 lbs)

```{r}
#1
smallwars <- select(starwars, height, mass, species)
mutate(smallwars, mass_lbs = mass * 2.2)
```

# Summarize data with `summarize()`

```{r}
summarize(starwars, mean_mass = mean(mass, na.rm = T))

##Group by species
summarize(starwars, mean_mass = mean(mass, na.rm = T), .by = species)

##Group by multiple groups
summarize(starwars, mean_mass = mean(mass, na.rm = T), .by = c(species, sex))
```

Only columns kept are the columns you group by and the columns you create.

`n()` is a helpful function that returns the number of rows in a group

```{r}
summarize(starwars, mean_mass = mean(mass, na.rm = T), 
          .by = species, 
          num_data = n())
```
